---
title: "Depth"
output: html_document
---

```{r setup, include = FALSE, message = FALSE, warning = FALSE}
library(ggplot2)
library(tidyverse)
library(showtext)
library(sysfonts)
library(workflowr)
library(readr)
library(wesanderson)

font_add_google("Lato", "lato")

showtext_auto()
```

# Sequencing coverage

Sequencing read alignments were processed using samtools. The input BAM file was first indexed, then coordinate-sorted and re-indexed to ensure compatibility with downstream analyses. Read depth was calculated for archaic samples at the specified pigmentation-related SNP panel using samtools depth including sites with zero coverage. 

```{r depth, include = FALSE, message = FALSE, warning = FALSE, echo = FALSE}
scladina <- read.table("data/Sclad_depth.txt", header = F) 
denny <- read.table("data/denny_depth.txt", header = F) 
mez1 <- read.table("data/mez1_depth.txt", header = F) 
hstadel <- read.table("data/hst_depth.txt", header = F) 
spy <- read.table("data/spy_depth.txt", header = F) 
lescottes <- read.table("data/lesCottes_depth.txt", header = F)
den5 <- read.table("data/den5_depth.txt", header = F)
mez2 <- read.table("data/mez2_depth.txt", header = F)
vindija87 <- read.table("data/vindija87_depth.txt", header = F)
chag8 <- read.table("data/Chagyrskaya8_depth.txt", header = F)
den25 <- read.table("data/den25_depth.txt", header = F)
den3 <- read.table("data/Denisova3_depth.txt", header = F)
vindija33 <- read.table("data/vindija33_depth.txt", header = F)
goyet <- read.table("data/goyet_depth.txt", header = F)

colnames(mez2) <- c("Chromosome", "Position", "Depth")
colnames(lescottes) <- c("Chromosome", "Position", "Depth")
colnames(hstadel) <- c("Chromosome", "Position", "Depth")
colnames(scladina) <- c("Chromosome", "Position", "Depth")
colnames(denny) <- c("Chromosome", "Position", "Depth")
colnames(mez1) <- c("Chromosome", "Position", "Depth")
colnames(spy) <- c("Chromosome", "Position", "Depth")
colnames(den5) <- c("Chromosome", "Position", "Depth")
colnames(chag8) <- c("Chromosome", "Position", "Depth")
colnames(vindija87) <- c("Chromosome", "Position", "Depth")
colnames(den25) <- c("Chromosome", "Position", "Depth")
colnames(den3) <- c("Chromosome", "Position", "Depth")
den3 <- den3 %>%
  mutate(Chromosome = as.numeric(substr(Chromosome, 3, nchar(Chromosome))))
colnames(vindija33) <- c("Chromosome", "Position", "Depth")
colnames(goyet) <- c("Chromosome", "Position", "Depth")

scladina$Sample <- "Scladina"
denny$Sample <- "Denisova 11 (Denny)"
mez1$Sample <- "Mez1"
mez2$Sample <- "Mez2"
hstadel$Sample <- "HST"
spy$Sample <- "Spy"
lescottes$Sample <- "Les Cottés Z4-1415"
den5$Sample <- "Denisova 5 (Altai Neanderthal)"
chag8$Sample <- "Chagyrskaya 8"
vindija87$Sample <- "Vindija 87"
den25$Sample <- "Denisova 25"
den3$Sample <- "Denisova 3"
vindija33$Sample <- "Vindija 33"
goyet$Sample <- "Goyet Q56-1"

depth_all <- bind_rows(scladina, denny, mez1, mez2, hstadel, 
                       spy, lescottes, den5, chag8, vindija87, 
                       den25, den3, vindija33, goyet)

depth_all <- depth_all %>%
  mutate(
    Chromosome = factor(Chromosome, levels = unique(Chromosome)),
    Sample = factor(Sample, levels = c("Scladina", "Denisova 11 (Denny)", "Mez1", 
                                       "Mez2", "HST", "Spy", "Les Cottés Z4-1415", 
                                       "Denisova 5 (Altai Neanderthal)", 
                                       "Chagyrskaya 8", "Vindija 87", "Denisova 25", 
                                       "Denisova 3", "Vindija 33", 
                                       "Goyet Q56-1"))
  )
```

```{r depth_viz, echo = FALSE}
wes_samples <- wes_palette("Zissou1", 14, type = "continuous")

ggplot(depth_all, aes(x = Position, y = Depth, color = Sample)) +
  geom_point(size = 2, alpha = 0.6) +
  scale_color_manual(values = wes_samples) +
  labs(
    title = "Sequencing Depth",
    subtitle = "EBI pigmentation-related loci",
    x = "Genomic Position",
    y = "Sequencing depth",
    color = "Sample"
  ) +
  theme_bw(base_size = 14) +
  theme_minimal(base_family = "lato") +
  theme(
    legend.position = "top",
    legend.background = element_blank(),
    panel.grid = element_blank(),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 10, hjust = 0.5)
  )
```


```{r depth_viz_log, echo = FALSE}
ggplot(depth_all, aes(x = Position, y = log(Depth), color = Sample)) +
  geom_point(size = 2, alpha = 0.6) +
  scale_color_manual(values = wes_samples) +
  labs(
    title = "Sequencing Depth (log-transformed)",
    subtitle = "EBI pigmentation-related loci",
    x = "Genomic Position",
    y = "Sequencing depth (log)",
    color = "Sample"
  ) +
  theme_bw(base_size = 14) +
  theme_minimal(base_family = "lato") +
  theme(
    legend.position = "top",
    legend.background = element_blank(),
    panel.grid = element_blank(),
    axis.text.x = element_blank(),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 10, hjust = 0.5)
  )
```


# Missingness

In order to examine patterns of missingness, I created a heatmap of where sequencing depth = 0. 


```{r missing_map}
depth_all <- depth_all %>%
  mutate(
    Position = as.factor(Position),
    Depth = as.numeric(Depth),
    Position_bin = round(Position / 500000),
    Depth_NA = ifelse(Depth > 0, log(Depth), NA)  # avoid -Inf
  )

ggplot(depth_all, aes(x = Position, y = Sample)) +
  geom_raster(aes(fill = Depth_NA)) +
  scale_fill_gradientn(
    colors = wes_palette("Zissou1", 100, type = "continuous"),
    na.value = "grey70",
    name = "log(Depth)"
  ) +
  theme_minimal(base_family = "lato") +
  theme(
    axis.text.x = element_blank(),
    axis.text.y = element_text(size = 8),
    panel.spacing = unit(0.5, "lines")
  ) 
```

Manhattan style plot 

```{r manhattan_depth}
depth_all <- depth_all %>%
  mutate(
    Position = as.numeric(as.character(Position)),
    Depth = as.numeric(as.character(Depth))
  )

depth_all <- depth_all %>%
  arrange(Chromosome, Position) %>%
  group_by(Chromosome) %>%
  mutate(chr_len = max(Position, na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(
    chr_offset = cumsum(lag(chr_len, default = 0)),
    Position_cum = Position + chr_offset,
    Depth_log = ifelse(Depth > 0, log10(Depth), NA)
  )

ggplot(depth_all, aes(x = Position_cum, y = Depth_log, color = Chromosome)) +
  geom_point(size = 0.8, alpha = 0.8) +
  scale_color_manual(
    values = rep(wes_palette("Zissou1", 5), length.out = length(unique(depth_all$Chromosome)))
  ) +
  theme_minimal(base_family = "lato") +
  labs(
    x = "Genomic Position",
    y = "log10(Depth)",
    color = "Chromosome"
  )

ggplot(depth_all, aes(x = Position_cum, y = Depth_log, color = Chromosome)) +
  geom_point(size = 0.6) +
  facet_wrap(~ Sample, ncol = 3) +
  scale_color_manual(
    values = rep(wes_palette("Zissou1", 5), length.out = length(unique(depth_all$Chromosome)))
  ) +
  theme_minimal(base_family = "lato") +
  theme(
    legend.position = "none",
    strip.text = element_text(face = "bold")
  ) +
  labs(
    x = "Genomic Position",
    y = "log10(Depth)"
  )
```




